\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-31 paper.qmd
Updated version for the discussion

- Summarise contributions
- Explain where it is expected or not expected to work, eg higher dimensional relationships
- Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views
- Predicting new observations in $k$-D
- Extending layouts beyond $k$-D, when 2D is clearly inadequate.
- Diagnostic app to explore differences in distances
- What might be useful enhancements

>>>>

<!-- - Summarise contributions -->
<!-- - Explain where it is expected or not expected to work, eg higher dimensional relationships -->
<!-- - Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views (need to add) -->
<!-- - Predicting new observations in $k$-D -->
<!-- - Extending layouts beyond $k$-D, when 2D is clearly inadequate. -->
<!-- - Diagnostic app to explore differences in distances (need to add) -->
<!-- - What might be useful enhancements -->

This study makes several important contributions to the field of NLDR. We have developed an algorithm to evaluate the most useful NLDR method and (hyper-)parameter choices for creating a reasonable \gD{} layout of high-dimensional data. Our objective is to fit a model for the \gD{} layout that preserves the relationships between neighboring points and turns it into a high-dimensional wireframe, which can be overlaid on the data and visualized using a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

The effectiveness of this approach is illustrated through various examples. For instance, the S-curve example demonstrates how the model accurately fits the points, capturing both local and global structures in high-dimensional space. Our simulation case study further, five Gaussian cluster example shows that while all observed NLDR methods preserve the global structure, only tSNE effectively maintains the local structure, highlighting the specific strengths and quirks of different methods.

Human behavior often shows a desire for more certainty and a tendency to prefer well-separated views. This emphasizes the importance of clear and distinct clusters. For example, in the UMAP layout of the **pbmc** dataset suggested by @chen2023, three distant, well-separated clusters are shown. However, our model reveals that these clusters are actually close to each other in \pD{}. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics.

Evaluating the error or unexplained variance is important for assessing how well the model fits the data. By examining the error for different numbers of bins, we found that tSNE with a perplexity of $30$ provides a reasonable representation for the **pbmc** dataset. Connecting the closest clusters with line segments in the fitted model further supports the preservation of neighborhood relationships.

The **digit: 1** example further illustrates the model's ability to accurately capture non-linear structures and provide additional information. Key findings include a twisted pattern that compresses the structure in some projections and long line segments that detect anomalies.

Predicting new observations in \kD{} is particularly valuable due to the limitations of some NLDR methods, like tSNE, which don't provide a straightforward method for prediction. As a result, our approach offers a solution that capable of generating predicted \kD{} embedding regardless of the NLDR method employed, effectively addressing this functional gap.

In conclusion, while our method effectively captures and represents high-dimensional data structures, further enhancements could involve introducing approaches to bind the data and indicate line segments beyond \gD{}. These improvements would help in creating a more accurate representation of the data when \gD{} layout is inadequate.

## Supplementary Materials

Code, and data for reproducing this paper are available at [https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm).

### 2024-05-31 paper.tex
Updated version for the discussion

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  Human behaviour, the desire to have more certainty, and a tendency to
  prefer the well-separated views
\item
  Predicting new observations in \(k\)-D
\item
  Extending layouts beyond \(k\)-D, when 2D is clearly inadequate.
\item
  Diagnostic app to explore differences in distances
\item
  What might be useful enhancements
\end{itemize}

>>>>

This study makes several important contributions to the field of NLDR.
We have developed an algorithm to evaluate the most useful NLDR method
and (hyper-)parameter choices for creating a reasonable \gD{} layout of
high-dimensional data. Our objective is to fit a model for the \gD{}
layout that preserves the relationships between neighboring points and
turns it into a high-dimensional wireframe, which can be overlaid on the
data and visualized using a tour. This approach is defined as
\emph{model-in-data-space}. Viewing a model in the data space is an
ideal way to examine the fit.

The effectiveness of this approach is illustrated through various
examples. For instance, the S-curve example demonstrates how the model
accurately fits the points, capturing both local and global structures
in high-dimensional space. Our simulation case study further, five
Gaussian cluster example shows that while all observed NLDR methods
preserve the global structure, only tSNE effectively maintains the local
structure, highlighting the specific strengths and quirks of different
methods.

Human behavior often shows a desire for more certainty and a tendency to
prefer well-separated views. This emphasizes the importance of clear and
distinct clusters. For example, in the UMAP layout of the \textbf{pbmc}
dataset suggested by \citet{chen2023}, three distant, well-separated
clusters are shown. However, our model reveals that these clusters are
actually close to each other in \pD{}. Additionally, the model discovers
non-uniform data distribution and non-linear structures within the
clusters that are not visible in the UMAP layout, demonstrating the
ability of our model in uncovering hidden data characteristics.

Evaluating the error or unexplained variance is important for assessing
how well the model fits the data. By examining the error for different
numbers of bins, we found that tSNE with a perplexity of \(30\) provides
a reasonable representation for the \textbf{pbmc} dataset. Connecting
the closest clusters with line segments in the fitted model further
supports the preservation of neighborhood relationships.

The \textbf{digit: 1} example further illustrates the model's ability to
accurately capture non-linear structures and provide additional
information. Key findings include a twisted pattern that compresses the
structure in some projections and long line segments that detect
anomalies.

Predicting new observations in \kD{} is particularly valuable due to the
limitations of some NLDR methods, like tSNE, which don't provide a
straightforward method for prediction. As a result, our approach offers
a solution that capable of generating predicted \kD{} embedding
regardless of the NLDR method employed, effectively addressing this
functional gap.

In conclusion, while our method effectively captures and represents
high-dimensional data structures, further enhancements could involve
introducing approaches to bind the data and indicate line segments
beyond \gD{}. These improvements would help in creating a more accurate
representation of the data when \gD{} layout is inadequate.

\section{Supplementary Materials}\label{supplementary-materials}

Code, and data for reproducing this paper are available at
\url{https://github.com/JayaniLakshika/paper-nldr-vis-algorithm}.
